---
title: "VeriNA3d: introduction and use cases"
author:
  - name: Diego Gallego
    affiliation:
    - Institute for Research in Biomedicine (IRB Barcelona), the Barcelona 
        Institute of Science and Technology, Spain.
    - Department of Biochemistry and Molecular Biomedicine, Faculty of Biology,
        University of Barcelona, Spain.
  - name: Leonardo Darre
    affiliation:
    - Institute for Research in Biomedicine (IRB Barcelona), the Barcelona 
        Institute of Science and Technology, Spain.
    - Functional Genomics Laboratory and Biomolecular Simulations Laboratory, 
        Institute Pasteur of Montevideo, Uruguay.
  - name: Pablo D. Dans
    affiliation:
    - Institute for Research in Biomedicine (IRB Barcelona), the Barcelona 
        Institute of Science and Technology, Spain.
  - name: Modesto Orozco
    affiliation:
    - Institute for Research in Biomedicine (IRB Barcelona), the Barcelona 
        Institute of Science and Technology, Spain.
    - Department of Biochemistry and Molecular Biomedicine, Faculty of Biology,
        University of Barcelona, Spain.
date: "`r Sys.Date()`"
output: BiocStyle::pdf_document
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction and use cases"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
    collapse   = TRUE,
    comment    = "#>",
    message    = FALSE,
    warning    = FALSE,
    fig.width  = 6,
    fig.height = 2,
    fig.wide   = TRUE
)

```
# Introduction: Structural Bioinformatics in R

The R language provides an excellent interface for statistical analysis, which
is also interesting from the point of view of structural data. This gap was 
filled in 2006 by the R package 
[bio3d](http://thegrantlab.org/bio3d/) [@Grant2006]. 
It was presented as a 
suite of tools to handle PDB formated structures, and trajectories. 
It integrates a variety of functions to analyse from 
sequence to 3D structure data (RMSD, NMA, PCA... see their 
[documentation](http://thegrantlab.org/bio3d/html/index.html) for details).
As far as we know, bio3d represented the only structural package for R until
now.

The R package presented in here, veriNA3d, does not replace bio3d at all.
Rather, it was developed on top of it to cover additional necessities. 
The only common tool integrated in both packages is a parser for mmCIF files
(see below). VeriNA3d is mainly intended (but not limited) to the analysis of 
Nucleic Acids. It integrates a higher level of abstraction than bio3d since it
also allows the analysis of datasets, in addition to analysis of single
structures. The functions in the package could be divided in the following 
blocks:

* Dataset level: Functions to get and analyse lists of pdb IDs. This includes
access to the [representative lists of RNA](
http://rna.bgsu.edu/rna3dhub/nrlist/)
by [@Leontis2012] and other analytical functions.

* Structure level: Functions to get data, parse mmCIF files and analyse them. 
This includes a wrapper of DSSR [@Lu2015] and a function to calculate
the \(\epsilon\)RMSD [@Bottaro2014].

* Plots: few functions to show the results of the previous analysis.

The complete list of functions can be found in the README.md file within the
package, also accessible on the gitlab 
[main page](http://mmb.irbbarcelona.org/gitlab/dgallego/veriNA3d).

\newpage
# Parsing mmCIF files

## Origin and standardization of mmCIF files

Atomic structural data of macromolecules has long been distributed in the PDB
file format. However, one of its main limitations is the column size for the 
coordinates data, which didn't allowed to save molecules with more than 99999
atoms, more than 62 chains or more than 9999 residues (in a chain).

Given that the Protein Data Bank is continously growing and accepting bigger
structures (e.g. a whole *E.coli* ribosome has over 140000 atoms - pdbID 4V4S),
an alternative file format became the standard: the mmCIF file format.

The mmCIFs are an evolution of the Crystallographic Information File (CIF), 
originally used for small molecule structures. It stands for 
**m**acro**m**olecular **CIF** file, and it has actually coexisted with the PDB
format since the 1997. However, since the PDB is easier to parse and such big
structures didn't populate the database at the time, most software has been
developed for the PDB format.

The PDB format was definetely frozen in 2014. However, it will still coexist
with the standard mmCIF format as long as all softwares evolve to accept
mmCIFs. Following this trend, the bio3d R package integrated a `read.cif` 
function in their version 2.3. At that time, we had already started the 
development of our own `cifParser` function. Given that the mmCIF format is
constantly evolving and that both functions take slightly different approaches,
we decided to offer our own version of it, which might provide an useful and
fast alternative for users working with mmCIF files.

## The CIF object

Parsing a particular file format often involves creating a new class of object.
In R, the principal [objects](http://adv-r.had.co.nz/OO-essentials.html) are 
called S3, S4 and RC. Our container for mmCIF data is an S4 object called 
**CIF**, in contrast with the S3 object (called **pdb**) in bio3d - it is 
worth noting that this difference does **not** affect the compatibility 
between the two packages (see below for details).

Since different mmCIF files usually have different sections of data 
(in addition to the coordinates), we carried out an analysis that checked 
which ones are always present in all mmCIF files (this included all mmCIF 
files in the Protein Data Bank in March 2018), and reached a list of 14
items:

* Atom\_site
* Atom\_sites
* Atom\_type
* Audit\_author
* Audit\_conform
* Chem\_comp
* Database\_2
* Entity
* Entry
* Exptl
* Pdbx\_database\_status
* Struct
* Struct\_asym
* Struct\_keywords 

The detailed description of each data sections can be found in the
mmCIF [main site](http://mmcif.wwpdb.org/).

The CIF object is created by the `cifParser` function and contains these 14
sections of data, which can be accessed with the CIF accessors. To see the
accessor functions run:

```{r ?cif_accessors}
library(veriNA3d)
?cif_accessors
```


To read a mmCIF file and access the coordinates data, use:

```{r cifParser(pdbID)}
## To parse a local mmCIF file:
# cif <- cifParser("your-file.cif")
## To download from PDB directly:
cif <- cifParser("1bau")
cif
## To see the coordinates:
coords <- cifAtom_site(cif)
head(coords)
```


## Bidirectional compatibility with bio3d

Using the `cifParser` will often be the first step to analyse a structure. 
However, if the analysis requires any of the bio3d functions, then a 
conversion should be done with the `cifAsPDB` function.

```{r cifAsPDB(cif)}
pdb <- cifAsPDB(cif)
pdb
```

It takes a CIF object and generates an equivalent pdb object (as used by all 
bio3d functions). In addition, all veriNA3d functions are prepared to accept 
as input either the CIF or pdb objects. Therefore, the compatibility between
the two packages is bidirectional.

\newpage
# Downloaded files

The `cifParser` and other functions download the mmCIF files when they are
not provided by the user. The core code that performs this operation is
`cifDownload`. 

`cifDownload` uses a temporal directory to save files, which can be found 
running `tempdir`. These files are saved accross the same R session, so no
file will be downloaded twice, but removed after the session is terminated.

In some cases, such as intensive use of the package or limited Internet 
connection, the user might want to keep the files accross different R sessions.
This functionality can be achieved with a simple process.

By default, `cifDownload` tries to find a directory called
"~/.veriNA3d_mmCIF_files/". Only when it does not find it, it will use a 
temporary directory. If you want to keep files across R sessions, just
create this new folder in your home directory and veriNA3d will start using it
to save all the downloaded mmCIF files. Note that the only way to recover the
space in your drive is to remove those files *manually*. 

Creating this directory can be easily done running the following command in R.

```{Create permanent directory}
dir.create("~/.veriNA3d_mmCIF_files/")
```

On the other hand, if you prefer to save your files in a specific directory
of your choice, you can create a symbolic link called .veriNA3d_mmCIF_files
pointing to your desired location, and it will also work. Instructions for this
particular task are not included, since they are system specific.

\newpage
# Getting data from Application Programming Interfaces (API)

Getting data is the first step of any pipeline, and parsing files is just one
of the many ways data can be accessed. Application Programming Interfaces (API)
are an intermediate point of access to a remote database. APIs offer the
users a series of *endpoints* or *calls*, which are just **links**. Thus, 
instead of dealing directly with the database with SQL or other language, the
user can just use the appropriate *call* to send a query to the API, and it 
will return the desired data.

In addition to parsing mmCIF files, veriNA3d also offers a series of functions
to send queries to different APIs. Since sending queries to remote APIs 
requires Internet access, the full functionality of veriNA3d might depend
on a good connection.

To see all the query functions, run:

```{r ?queryFunctions}
?queryFunctions
```

IMPORTANT NOTE: 
The APIs accessed by veriNA3d are free of use with no limit of queries
per user. However, this could change if the users of the APIs use them 
irresponsibly. Servers could eventually fall down if they receive more 
*calls* than they can actually process. To avoid that, veriNA3d actually
saves in memory the result of any query, and any time that you use the same
query again, it will take the cached result. To see this effect, run this test:

```{r "queryTechnique(pdbID, verbose=TRUE)"}
## Run a query for the first time, which will access the API
tech <- queryTechnique("4KQX", verbose=TRUE)

## Run the same query for the second time, which will get it from memory
tech <- queryTechnique("4KQX", verbose=TRUE)

## See result
tech
```

However, this is only saved across the current session of R, and any script
that uses the query functions will send them to the API every time it is run.
VeriNA3d does not guarantee the correct service of the APIs and it does not
monitor any of your processes. However, the API providers 
can known which IP address is querying their services at any time. To avoid 
overloading the servers, please make a responsible use (e.g. save locally the 
data that you use frequently).

## Querying the EMBL-EBI REST API

An invaluable resource for structural & computational biologists is the 
[PDBe REST API](https://www.ebi.ac.uk/pdbe/pdbe-rest-api) [@Velankar2015].
Around this technology, the package includes the following set of functions:

* queryAuthors: List of authors
* queryReldate: Release date
* queryDepdate: Deposition date
* queryRevdate: Revision date
* queryDescription: Author description
* queryEntities: Entitity information
* queryFormats: File formats for the given ID
* queryModres: Modified residues
* queryLigands: Ligands in structure
* queryOrgLigands: Ligands in structure (substracting ions)
* queryResol: Resolution (if applicable)
* queryTechnique: Experimental Technique
* queryStatus: Released/Obsolete and related status information

The list of functions is **intentionally limited** in comparison with the 
dozens of *endpoints* of the REST API. Integrating them all would needlessly
increase the total amount of functions of the package. Moreover, the API
might offer more and more *endpoints* with the time, and trying to keep them
all would make the manteinance of this package more difficult. To allow the 
users to access their desired endpoints, an alternative method is provided.

The core of all the query functions is `queryAPI`, which
integrates all the error-handling and cache functionalities. With the 
`queryAPI` function, any user can design their own queries, with a simple
process. Herein a couple of examples.

### Example 1

This snapshot shows the REST API website and a *call* that is not implemented
in veriNA3d:

\begin{center}
\includegraphics[width=6in]{PDBe_REST_API.png}
\end{center}

The link of this *endpoint* is:
https://www.ebi.ac.uk/pdbe/api/pdb/compound/summary/ATP

The `queryAPI` function can understand and send this query using the arguments
'ID', 'API', 'string1', and 'string2' properly:

```{r "Example1: queryAPI(ID, API, string1, string2)"}
atpsummary <- queryAPI(ID="ATP", API="ebi", 
                        string1="pdb/compound/summary/", string2="")
str(atpsummary$ATP)
```

* The common root in all the REST API *endpoints* is 
"https://www.ebi.ac.uk/pdbe/api/", which is internally managed by the function
by using *API="ebi"*.
* The *string1="pdb/compound/summary/"* indicates everything that comes after
the root and before the ID. 
* The *ID="ATP"* obviously represents the desired structure, either a
4 character string for a pdbID or <= 3 character string for compounds.
* The *string2=""* is also a necessary argument that reflects that nothing
comes after the ID.

### Example 2

This snapshot shows a second *call* not implemented in veriNA3d:

\begin{center}
\includegraphics[width=6in]{PDBe_REST_API2.png}
\end{center}

The link of this *endpoint* is: 
https://www.ebi.ac.uk/pdbe/api/pisa/noofinterfaces/3gcb/0

The proper call with `queryAPI` would be:

```{r "Example2: queryAPI(ID, API, string1, string2)"}
PISAsummary <- queryAPI(ID="3gcb", API="ebi", 
                        string1="pisa/noofinterfaces/", string2="0")
str(PISAsummary$"3gcb")
```

This second example shows a case in which the "string2" argument is necessary.
If you are unsure about the real link that is actually being constructed,
you can always use *verbose=TRUE* to see it printed.

\newpage
# The \(\epsilon\)RMSD to compare structures

A new interesting metric to compare Nucleic Acid structures is the 
\(\epsilon\)RMSD [@Bottaro2014], currently available in the BaRNAba python
package [@Bottaro2018]. The metric was implented in the `eRMSD` function
and completely reproduces BaRNAba results for the structures tested.

The following example shows how to get the \(\epsilon\)RMSD between two 
models of the same structure:

```{r "eRMSD(structure1, structure2)"}
## Parse cif file
cif <- cifParser("2d18")
## Select a couple of models
model1 <- selectModel(cif=cif, model=1)
model3 <- selectModel(cif=cif, model=3)

## Calculate the eRMSD
eRMSD(cif1=model1, cif2=model3)

## The RMSD can also be calculated easily
RMSD(cif1=model1, cif2=model3)

```

\newpage
# Generate substructures

In many cases you might be interested on a particular region of a structure
(e.g. a peptide from a complex, or a ligand and its binding site). For a 
given structure, `trimSphere` can generate a smaller pdb object and save it 
to a PDB file if desired. The region of interest can be selected by using 
the chain identifier and the residue index, or with the `atom.select` function
from bio3d (which in turn allows you to select regions of the structure in a 
variety of ways). In addition to the region of interest, the function can also
include the surrounding region by seting a cutoff distance.

### Example 1

```{r "trimsphere(structure, chain)"}
## Parse human ribosome - takes around 12 seconds in R-3.5
cif <- cifParser("6ek0")

## Query entities and check them
ent <- queryEntities("6ek0")
head(ent[, c("entity_id", "molecule_name", "in_chains")])

## Generate a smaller pdb with the 60S ribosomal protein L8
chain <- "LA"
protL8 <- trimSphere(cif, chain=chain, cutoff=0)
protL8

## The same command with the argument file would save it directly:
trimSphere(cif, chain=chain, cutoff=0, file="output.pdb")
```

The output.pdb file can then be visualized using your favourite external
molecular viewer. The following picture was taken using VMD software
[@Humphrey1996].

\begin{center}
\includegraphics[width=6in]{6ek0_L9.png}
\end{center}

### Example 2

To get the desired region of interest and its sorroundings, let's see a 
second example using the same structure:

```{r "trimsphere(structure, selection)"}
## Load bio3d library
library(bio3d)

## Get pdb object from CIF
pdb <- cifAsPDB(cif)

## Get list of ligands in the human ribosome 6EK0
queryLigands("6ek0", onlyligands=T)

## Get the atomic index for a desired ligand
HMTligand_inds <- pdb$atom$eleno[which(pdb$atom$resid == "HMT")]

## Use bio3d function to select the ligand using its atom indices
sel <- atom.select(pdb, eleno=HMTligand_inds)

## Get substructure and sorroundings at 10 Angstroms
HTMligand <- trimSphere(pdb, sel=sel, cutoff=5)

## And generate file to visualize it
trimSphere(pdb, sel=sel, file="output2.pdb", cutoff=5)
```

\begin{center}
\includegraphics[width=6in]{6ek0_HMT.png}
\end{center}


### Example 3

A third useful example would be the generation of a pdb with the
interacting region between two molecules in the structure. To achieve the
goal, veriNA3d also counts with the `findBindingSite` function, as shown
below:

```{r "trimsphere(structure, selection2)"}
## Parse another pdb for this example
pdb <- cifAsPDB("1nyb")

## Find region of interaction between RNA and protein
data <- findBindingSite(pdb, select="RNA", byres=TRUE)

## Get atom indices from interacting region molecules
eleno <- append(data$eleno_A, data$eleno_B)

## Select using bio3d
sel <- atom.select(pdb, eleno=eleno)

## Get substructure
trimSphere(pdb, sel=sel, file="interacting_site.pdb", verbose=FALSE)
```

\begin{center}
\includegraphics[width=6in]{1nyb_interacting_site.png}
\end{center}

\newpage
# Calculate phosphate pair-wise distances

Atomic distances are calculated internally in the `trimSphere` function, a 
resource that is also made available to the user through the functions
`measureEntityDist` (returns all the distances between the atoms 
of two entities) and `measureElenoDist` (returns all the distances between two
sets of atoms).

Herein it is shown how to calculate the pair-wise distances between all the
phosphate atoms of an RNA structure, which could be easily adapted for alpha
carbons or other elements.

```{r measureElenoDist(structure, refeleno, eleno, cutoff)}
## Parse pdb
pdb <- cifAsPDB("1nyb")

## Select P atoms by element number (eleno)
ind <- which(pdb$atom$elety == "P")
eleno <- pdb$atom$eleno[ind]

## Count number of phosphates
total <- length(eleno)

## Execute function to measure the distances
P_distances <- measureElenoDist(pdb=pdb, refeleno=eleno, eleno=eleno, 
                                n=total, cutoff=100)
```

\newpage
# Manage Nucleic Acid datasets

VeriNA3d also provides a set of tools to work at a higher level of abstraction,
which allow the user to manage datasets of structures at once. Such approaches
are often found in the literature in a variety of studies, either
with descriptive [@Murray2003, @Darre2016], validation [@Jain2015] or 
modeling [@Wang2011, @Wang2015] purposes.
However, as far as we know, no software with dataset analysis
purposes has ever been freely distributed before.

## Non-redundant lists of RNA

VeriNA3d integrates
the function `getRNAList`, which returns the desired non-redundant list
of RNA structures, as found in [@Leontis2012]
[website](http://rna.bgsu.edu/rna3dhub/nrlist/). 
Here it is shown an example of how to make a non-redundant dataset of 
protein-RNA complexes with resolutions better than 2 Angstroms.

```{r getRNAList(release, threshold)}
## Get non-redundant list from Leontis website
rnalist <- getRNAList(release="3.47", threshold="2.0A")
head(rnalist)
```

The previous object is a data-frame with three columns:

* Equivalence class
* Representative structure
* Members of the Equivalence class

Note that the output is formated according with Leontis&Zirbel nomenclature
(AAAA|M|C), where "AAAA" is the PDB accession code, "M" is the model and "C" 
is the Chain.

The representative structures might not necessarily have the desired properties
(e.g. being a protein-RNA complex). To get only representatives which are
complexed with a protein, all the members of the class should be checked,
which can be done with `getAltRepres`.

```{r 'getAltRepres(rnalist, type="protRNA")'}
## Set progressbar=TRUE to see the progress
protrna <- getAltRepres(rnalist=rnalist, type="protRNA", progressbar=FALSE)
head(protrna)
```

Many equivalence classes will lack a proper representative, so a NA value is
returned. To get the definite list in a more friendly format and removing
missing data, use `represAsDataFrame`.

```{r represAsDataFrame(nrlist)}
nrlist <- represAsDataFrame(protrna)
head(nrlist)
```

## Queries

When analysing a dataset, a particular piece of data might be useful, such
as the file formats available or if a particular ligand is present or not in
all the pdb files. A handy function to approach this need is `applyToPDB`.
Here it's shown how to get the structures in the previous list that contain
Mg ions:

```{r applyToPDB(listpdb=nrlist, FUN=hasHetAtm, hetAtms="MG")}
## Set progressbar=TRUE to see the progress
nrlist_mg <- applyToPDB(FUN=hasHetAtm, listpdb=nrlist$pdb, 
                        hetAtms="MG", progressbar=FALSE)
nrlist <- cbind(nrlist, Mg=nrlist_mg[, 2])
head(nrlist)

## To see only the structures containing Mg use
head(nrlist[nrlist$Mg == TRUE, ])
```

In the example, the argument hetAtms is passed to the hasHetAtm function.

## Analysing datasets

When you have your desired list of PDB IDs, the function `pipeNucData` and 
`pipeProtNucData` can extract multiple structural parameters for all the 
dataset.

```{r pipeNucData(pdblist, chainlist, cores)}
## After the download is finished, this dataset is analysed in less than 2 min
## (single core, intel i5 2.3Ghz). Set progressbar=TRUE to see the progress.
ntinfo <- pipeNucData(pdbID=nrlist$pdb, 
                        model=nrlist$model,
                        chain=nrlist$chain, 
                        progressbar=FALSE, cores=2)
str(ntinfo)
```

In this case, `pipeNucData` is downloading each structure from internet, but
you can also download them manually before execution, save them in a directory
and provide it to the function. When reading files locally, the function is
obviously faster. The file format can either be pdb or cif, being the cif files
the recommended option (they can also be read when compressed in .gz). To make
it work with local files, the extension should also be provided
(e.g. ".cif.gz", ".cif" or ".pdb").

```{r pipeNucData(pdblist, chainlist, cores, path, extension), eval=FALSE}
ntinfo <- pipeNucData(pdbID=nrlist$pdb, 
                        model=nrlist$model,
                        chain=nrlist$chain, 
                        progressbar=FALSE, cores=2, 
                        path="/your/path/to/the/dataset/", extension=".cif.gz")
```

The second dataset analysis available in the dataset is designed to study
the interactions between proteins and nucleic acids, and would be executed as
follows:


```{r pipeProtNucData(pdblist, chainlist, cores)}
## After the download is finished, this dataset is analysed in less than 1 min
## (single core, intel i5 2.3Ghz). Set progressbar=TRUE to see the progress.
aantinfo <- pipeProtNucData(pdbID=nrlist$pdb, 
                            model=nrlist$model,
                            chain=nrlist$chain, 
                            progressbar=FALSE, cores=2)
str(aantinfo)
```

\newpage
# The `dssr` wrapper: getting the base pairs

The DSSR software (Disecting the Spatial Structure of RNA) [@Lu2015] 
represents an unvaluable resource to handle RNA structures. Some of the 
functions of veriNA3d overlap with the functionalities of DSSR, and both 
applications provide unique different features. We implement a wrapper to
execute DSSR directly from R and get the best of both worlds in one place.

Note that installing veriNA3d does not automatically install DSSR, since
we don't redistribute third-party software. Before any user can use our 
wrapper, the `dssr` function, DSSR should be installed separately. To
address this installation we redirect you to the 
[DSSR manual](http://docs.x3dna.org/dssr-manual.pdf), where anyone can find
the specific instructions for their system. Once DSSR is installed and working 
in your computer, you will also be able to use it with our wrapper. If the
DSSR executable (named x3dna-dssr) is in your path, `dssr` will find it
automatically. If the wrapper does not find it, you can still use it
specifying the absolute path to the executable with the argument exefile.
Find more information running `?dssr`.

One of the DSSR capabilities that users might be interested in is the detection
and classification of base pairs. The following code shows a simple example.
The output of the `dssr` wrapper is an object got from the json DSSR output.
From R, json objects are parsed in the form of a tree of lists, with different
types of information. Most of the interesting data is under the list models,
sublist parameters, as shown herein.


```{r dssr(pdbID)}
## Execute dssr, the wrapper will donwload the (mmCIF) file if necessary
rna <- dssr("1bau")

## The contents of the `rna` object can be seen with
names(rna)

## Then, the contents of the list model inside `rna`
names(rna$model)

## And, inside the sublist parameters
names(rna$models$parameters)

## The number 1 herein selects the model of the structure, and is needed even
## in XRAY structures.
class(rna$models$parameters$pairs[[1]])

## Finally, to get the data about base pairs, use:
rna$models$parameters$pairs[[1]][, c("index", "nt1", "nt2", "LW")]

```

\newpage
# References
